---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Libraries used: dplyr, ggplot2

```{r include=FALSE}
library(dplyr)
library(ggplot2)
setwd("C:/Users/Gary/Documents/R Files/Reproducible/RepData_PeerAssessment1")
```
Data was loded into a single dataframe named "df"
```{r}
df <- read.csv("activity.csv")
str(df)
```

## What is mean total number of steps taken per day?
Create a vector of all dates present in the dataframe
```{r}
Dates <- as.Date(strftime(df$date, "%Y-%m-%d"))
allDates <- seq(from = min(Dates), to = max(Dates), by = "day")
str(allDates)
```

Use sapply to get a count of steps for each day in the date vector and then 
add them to new data frame

```{r}
step.count <- sapply(allDates, FUN = function(X) sum(df$steps[Dates == X]))
totals <- data.frame(day = allDates, total_steps = step.count)
str(totals)
```
Use the base plotting system to create a histogram (quantiles)
```{r}
hist(totals$total_steps, breaks=5, xlab="Total Steps", main="Histogram of Total Steps")
```

Calculate the mean and median of the total steps taken per day and print them to screen
```{r}
step_mean <- mean(totals$total_steps, na.rm = TRUE)
print(paste(c("Average number of steps per day", step_mean), collapse = " : "))

step_median <- median(totals$total_steps, na.rm = TRUE)
print(paste(c("Median number of steps per day", step_median), collapse = " : "))
```

## What is the average daily activity pattern?
Calculate the average number of steps per 5 minutes for each day and place into a data frame
```{r}
average_steps <- sapply(allDates, FUN = function(X) mean(df$steps[Dates == X]))
average_totals <- data.frame(day = allDates, avg_steps = average_steps)
str(average_totals)
```

Create a line plot of the average number of steps taken per five minutes by day

```{r}
plot(average_totals$avg_steps ~ average_totals$day, average_totals, xaxt = "n", type = "l", xlab="Date", ylab="Average Steps", main = "Avgerage Number of Steps per 5 Minute Interval")
axis(1, average_totals$day, format(average_totals$day, "%b %d"), cex.axis = .7)

```

## Inputing missing values
Report on the total number of missing records
```{r}
total_nas <- sum(is.na(df$steps))
print(paste(c("Total number of missing values", total_nas), collapse = " : "))
```

Calculate the average number of steps per 5 minute interval
```{r}
avg_steps <- sum(df$steps, na.rm = TRUE)/nrow(df)
print(paste(c("Average number of steps per 5 minute interval", avg_steps), collapse = " : "))
```

Replace NA values with the average number of steps
```{r}
df$steps[is.na(df$steps)] <- avg_steps
total_nas <- sum(is.na(df$steps))
print(paste(c("Total number of missing values", total_nas), collapse = " : "))
str(df)
```

### Replot the histogram after missing values have been replaced

Recalculate the daily total
```{r}
step.count <- sapply(allDates, FUN = function(X) sum(df$steps[Dates == X]))
totals <- data.frame(day = allDates, total_steps = step.count)
```

Use the base plotting system to create a histogram (quantiles)
```{r}
hist(totals$total_steps, breaks=5, xlab="Total Steps", main="Histogram of Total Steps")
```

Calculate the mean and median of the total steps taken per day and print them to screen
```{r}
step_mean <- mean(totals$total_steps, na.rm = TRUE)
print(paste(c("Average number of steps per day", step_mean), collapse = " : "))

step_median <- median(totals$total_steps, na.rm = TRUE)
print(paste(c("Median number of steps per day", step_median), collapse = " : "))
```
## Are there differences in activity patterns between weekdays and weekends?

Get a list of dates and their correcponding day of the week
```{r}
wd <- weekdays(as.Date(df$date))
str(wd)
```

Create a weekday/weekend vector
```{r}
ww <- sapply(wd, function(x){if(x == "Saturday" || x == "Sunday") "weekend" else "weekday"})
str(ww)
```

Bind it to the dataframe
```{r}
df <- mutate(df, day_type = ww)
str(df)
```


Create a panel plot using ggplot
```{r}
p <- ggplot(df, aes(x=date, y=steps, group=1))+geom_line(color="aquamarine4")+facet_wrap(~day_type, nrow=2)
print(p)
```